---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}
data <- read.csv("activity.csv")
str(data)
```



## What is mean total number of steps taken per day?

```{r}
stepsperday <- tapply(data$steps, data$date, sum)
hist(stepsperday, breaks = 10)
```  

Mean of the total number of steps taken per day:
```{r}
mean1 <- mean(stepsperday, na.rm = T)
mean1
```
Median of the total number of steps taken per day:
```{r}
median1 <- median(stepsperday, na.rm = T)
median1
```



## What is the average daily activity pattern?  

1. Make a time series plotof the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)  

```{r, message=FALSE}
library(dplyr)
```

```{r}

avgsteps<-data %>% select(steps,interval) %>% group_by(interval) %>% summarise(average_steps = mean(steps, na.rm = T)) 

plot(avgsteps$interval,avgsteps$average_steps, type = "l", xlab = "Interval", ylab = " Average Daily Steps", main = "Average Daily Activity Pattern")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps? 

```{r}
filter(avgsteps, average_steps == max(average_steps))
```


## Imputing missing values


```{r, message=FALSE}
library(dplyr)
```

Checking how many missing values are there:  
```{r}
colSums(is.na(data))

```
Imputing missing values with mean for the corresponding day and save it as data2
```{r}
data2<-data %>% group_by(interval) %>%
mutate(steps=ifelse(is.na(steps),mean(steps,na.rm=TRUE),steps))
```

Checking for missign values now:

```{r}
colSums(is.na(data2))

```

Histogram of total number of steps per day

```{r}
stepsperday2 <- tapply(data2$steps, data2$date, sum)
hist(stepsperday2, breaks = 10)
```
  
Mean of the total number of steps taken per day:
```{r}
mean2 <- mean(stepsperday2, na.rm = T)
mean2
```
Median of the total number of steps taken per day:
```{r}
median2 <- median(stepsperday2, na.rm = T)
median2
```

Affects on the mean and median after imputing missing values:  
  
Mean before imputing: `r mean1`  

Mean after imputing: `r mean2`

Median before imputing: `r median1`  

Median after imputing: `r median2`  

We can see that the mean remained exactly the same which is as expected because we used the mean by interval to replace the missing values. On the other hand theres a very small change in the median in fact mean and median are same now.  



## Are there differences in activity patterns between weekdays and weekends?  

```{r}
library(ggplot2)
```

Creating a new factor variable in the dataset with two levels – “weekday” and “weekend”  
```{r}
data2$date <- as.Date(data2$date)
data2$week <- ifelse(weekdays(data2$date) %in% c("Saturday", "Sunday"), "weekend", "weekday")
```

Making a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).  
 
```{r}
avgsteps2<-data2 %>% select(steps,interval, week) %>% group_by(interval,week) %>% summarise(average_steps = mean(steps, na.rm = T)) 

p <- ggplot(avgsteps2, aes(x=interval, y=average_steps)) +
  geom_line() + 
    facet_grid(week~.)
p

```


